# Wrangling and EDA With Time Series Data

## Introduction

One of the first few steps of any data analysis project is wrangling with data and performing exploratory analysis (EDA).
The former involves cleaning and restructuring raw data into a more organized and usable form, while the latter comprises creating plots, graphs, summary statistics, and using them to guide further analysis.

In this chapter, we cover the special characteristics of data wrangling and EDA for time series data.
It will also double as an introduction to time series modeling packages in R and Python.
As we will see in @sec-eda-packages, the software ecosystem for time series analysis in both R and Python is quite extensive.
In this course, we will primarily make use of the `fpp3` package associated with the textbook @hyndman2018forecasting.

```{r}
#| message: TRUE
#| warning: FALSE
library(fpp3)
# help(package="fpp3")
```

Running the above script, we see that loading `fpp3` automatically loads `tsibble`, `tsibbledata`, `feasts` and `fable`, along with several `tidyverse` packages.
It also comes preloaded with a number of datasets that are useful for pedagogical purposes.

## Tsibbles

A time series comprises a series of measurements along with information about when those measurements were taken (the time index).
The `tsibble` package implements a convenient data structure, the tsibble, for storing time series data.
To illustrate how they work, we create a simple example as follows.

```{r}
#| message: TRUE
#| warning: FALSE
#| output: FALSE
example1 <- tsibble(
  year = 2015:2019,               # <1>
  y = c(123, 39, 78, 52, 110),    # <2>
  index = year                    # <3>
)

str(example1)
# output:
# tbl_ts [5 × 2] (S3: tbl_ts/tbl_df/tbl/data.frame)
# ...
```

1.  Creates a vector of years.
2.  `y` is the vector of values in the time series.
3.  The index column refers back to the `year` column that was just created.

Applying `str()` to the created object shows that it has class `tbl_ts` (tsibble). 
It also shows that it also inherits from `tbl_df` (tibble) and base R `data.frame`.
This means that all `dplyr` verbs (such as `filter`, `select`, and `mutate`) work as per normal.
However, there are now 3 types of columns in a tsibble:

1.  *Measurement* variables. Note that there could be more than one type of measurement at each time point.
2.  The *index* variable: a **single** column denoting the time point of each measurement.
3.  *Key* variables: a set of columns whose unique combinations define a single time series.

We discuss the second and third types in more detail.

### The index variable

The index column is what introduces the temporal component of the data. 
It is what associates the measurements in each row with a specific time point. 
The time point should be from a time class in R. 
The time points could correspond to yearly, monthly, weekly, daily and sub-daily intervals.
Consider the examples below, noting that the annotation "[1Y]", "[1M]", etc. tells us the interval between observations of each time series.

::: panel-tabset
## Annual data

```{r}
aus_airpassengers
```

Total annual air passengers (in millions) including domestic and international
aircraft passengers of air carriers registered in Australia.

## Monthly data

```{r}
us_employment
```

US employment data from January 1939 to June 2019. Each 'Series_ID' represents different sectors of the economy.

## Quarterly data

```{r}
aus_arrivals
```

Quarterly international arrivals to Australia from Japan, New Zealand, UK and
the US. 1981Q1 - 2012Q3.

## Weekly data

```{r}
us_gasoline
```

Million barrels per day, beginning Week 6, 1991, ending Week 3, 2017.
:::

The easiest way to create such a column is to use one of the convenience
functions from the `lubridate` package in R.

| Frequency | Function                   |
|-----------|----------------------------|
| Annual    | Use integers in R          |
| Quarterly | `yearquarter()`            |
| Monthly   | `yearmonth()`              |
| Weekly    | `yearweek()`               |
| Daily     | `as_date(),ymd()`          |
| Sub-daily | `as_datetime(), ymd_hms()` |

: lubridate functions for index column

### Key variables

A tsibble can store multiple time series in a single tsibble. 
It makes sense to do this when the timepoints overlap, and we wish to compare and contrast different time series.
Consider the dataset `us_employment`, containing US employment data from January 1939 to June 2019.

```{r}
#| message: TRUE
#| warning: FALSE
#| output: FALSE
us_employment
# Output:
# A tsibble: 143,412 x 4 [1M]
# # Key:       Series_ID [148]
#       Month Series_ID     Title         Employed
#       <mth> <chr>         <chr>            <dbl>
```

The printed output states that the tsibble's key is determined by a single column: `Series_ID`
Inspecting the metadata by running `help(us_employment)` tells us that each 'Series_ID' represents different sectors of the economy.
There can of course be multiple key columns, such as in `olympic_running`, which records the fastest running time for each Olympic event.

```{r}
#| message: TRUE
#| warning: FALSE
#| output: FALSE
olympic_running
# Output:
# # A tsibble: 312 x 4 [4Y]
# # Key:       Length, Sex [14]
#     Year Length Sex    Time
#    <int>  <int> <chr> <dbl>
```

The unique combinations of key columns is the number of time series in the
tsibble.


::: callout-note
-   There could be multiple measurement columns and multiple key columns in a
    tsibble.
-   There could be columns that are neither measurement nor key columns in a
    tsibble.
-   There must be exactly one index column in a tsibble.
:::

### Creating tsibbles

Time series data in the wild does not automatically occur as tsibbles.
They are instead stored as csv files or in other formats, and hence have to be converted to tsibble format.
The function `as_tsibble()` converts a tibble object into a tsibble.
It takes two arguments, `index` and `key`, which are used to specify the index and key columns respectively.
This often has to be preceded by several `dplyr` function calls on the tibble to make sure it has the appropriate format.
For an example on how to do this, refer to [Chapter 2.1](https://otexts.com/fpp3/tsibbles.html) in @hyndman2018forecasting.

## Time plots

The time plot is the most natural first thing to do with a time series. 
It plots the measurements on the y-axis, time on the x-axis and joins the points together with a line.
We have already seen a number of time plots in the previous section.

Generating a time plot from a tsibble is incredibly easy via `autoplot()`.

```{r}
#| fig-align: center
#| fig-cap: Air transport passengers on Australian air carriers.
#| label: fig-eda-timeplot
aus_airpassengers |> 
    autoplot(.vars = Passengers) + 
    geom_point() + 
    labs(y = "Passengers (millions)", 
         x = "Year")
```

There is a general upward trend of passengers.
There were two noticeable dips---one in 1989 and one in 2009 (what happened?).
It is arguable that the trend appears piecewise linear, with a more positive gradient after 1990.

When the tsibble contains multiple time series, `autoplot()` makes a line plot for each of them, i.e. the number of lines is the number of unique keys.
This can easily lead to overplotting, so one is advised to first `filter` for only the most relevant keys before plotting. 

## Common time series patterns

In @sec-examples-tsdata, we saw a diverse range of time series behaviors, as illustrated by their time plots.
There are nevertheless several patterns recur

## Seasonal plots


## Scatter plots

## Lag plots

## Summary statistics

## Where to find time series data


## Package ecosystem {#sec-eda-packages}

Both R and Python have 
